A Quantitative Genetic Linked Selection Model fit to the Negative Selection Process in Humans

### 2023-07-06 manuscript/manuscript.tex
ms changes

Overall, across our CADD-based models and all populations, our mutation rate
estimates are roughly concordant with pedigree-based estimates (Figure
\ref{fig:figure-4}A). Our full-track CADD 6\% model estimates a mutation rate
of $\mu = 1.46 \times 10^{-8}$ for YRI, $1.53 \times 10^{-8}$ for CEU, and $1.5
\times 10^{-8}$ for CHB reference samples (Supplementary Table
\ref{supp:tbl-r2}). Mutation rate estimates for our feature-based annotation
models are unrealistically higher. We believe this could be caused by
per-basepair heterogeneity in selective effects within each of these features.
For example, mutations in coding sequences are known to have highly
heterogeneous effects depending on whether the result is synonymous,
non-synonymous, or mis-sense. If some fraction of the impact of negative
selection is due to strong selection at a few sites, sparser annotation models
like these cannot isolate these effects like CADD-based models. Instead, XXXHERE

reflects a general problem
with these more interpretable, but sparser feature-based tracks: heterogeneity
in selective effects

However, we noticed that estimates of mutation rate is very sensitive to the
choice of annotation model. Between CADD 6\% and 6\% models, the

@@

Given

>>>>

Across all populations, our mutation rate estimates from CADD-based models are
roughly concordant with pedigree-based estimates (Figure \ref{fig:figure-4}A).
Our full-track CADD 6\% model estimates the mutation rate as $\widehat{\mu} =
1.46 \times 10^{-8}$ for YRI, $1.53 \times 10^{-8}$ for CEU, and $1.5 \times
10^{-8}$ for CHB reference samples (Supplementary Table \ref{supp:tbl-r2}). As
expected, the sparse-track CADD model mutation rate estimates are identical
between the B' and classic BGS methods (Figure \ref{fig:figure-4}A top row).

However, mutation rate estimates for feature-based annotation models show
several issues. First, mutation rate estimates under using our implementation
of classic BGS theory are an order of magnitude below the expected range
(Figure \ref{fig:figure-4}A top row). We observe similar behavior when we use
the classic BGS model to fit full-coverage annotation models (Figure
\ref{fig:figure-4}A bottom row). This behavior is consistent with the fact that
classic BGS theory cannot fit weak selection, and to accommodate features under
weak constrain (e.g. introns, UTRs, and the ``other" feature) it must severally
underestimate the mutation rate.

Second, we noticed that across all populations and sparse/full tracks, the CADD
6\% model consistently led to slightly higher mutation rates than the CADD 8\%
model (Figure \ref{fig:figure-4}A bottom row; Supplementary Table
\ref{supp:tbl-r2}). This pattern was observed in Murphy et al. too (2023;
Appendix 1, Figure 16). This behavior is consistent with a non-identifiability
issue between a slightly higher per-basepair mutation rate and annotation
tracks that contain more conserved sequence. This is expected from theory,
since both classic BGS and SC16 models only depend on mutation rate through the
compound parameter $\mu L$ where $L$ is the length of the conserved segment.
Even though our method is much more robust to including non-conserved regions
like introns, we still observe this non-identifiability issue.

Finally, we note that mutation rate estimates in the Feature Priority model are
unreasonably high ($\widehat{\mu} \approx 4.2 \times 10^{-8}$), reminiscent of
the high mutation rate estimates found under McVicker et al.'s
(\citeyear{McVicker2009-ax}) model. While both our and Murphy et al.'s CADD and
PhastCons-based models alleviate this issue, it is worth discussing why this
could occur. We gain some insight from comparing the estimated mutation rates
between our Feature and PhastCons Priority annotation models, which both
contain the exact same number of basepairs, but these are assigned to different
feature classes based on the priority of overlapping features. That one of
these models is our best-fitting model, and the other our worst indicates a
large role in per-basepair heterogeneity in selective effects in determining
the genome-wide linked selection signal. CADD-based models fit well in part due
to their fine-scale resolution of selective effects across the genome.

@@

Given these the sensitivity of mutation rate estimates and the strong-selection
non-identifiability issue, we next sought to assess the estimation accuracy of
our composite likelihood method using forward simulations. In order to have
enough megabase-scale data to apply our method, we independently evolved the
first five human chromosomes. Due to the immense computational expense of
conducting forward simulations at this scale over numerous parameter
combinations, we used fixed mutation and selection parameters in putatively
conserved regions (see Methods XXX). We note three important findings from
these simulations.

First, the linked selection signal, as measured by the $R^2$ between predicted
and simulated megabase-scale diversity, increases with the intensity of
selection against new deleterious mutations and mutation rate (Figure
\ref{fig:figure-4}B, top row). This is consistent across both classic BGS and
our B' method; the only exception is the classic BGS model run with a
full-coverage track for the reasons described above. However, if new mutations
are highly deleterious ($s=0.05$), $R^2$ is reduced. This is consistent with
very strong selection having less localized effects and acting more like a
genome-wide faster rate of drift.

Second, both classic BGS theory and our B' method accurately infer the average
selection coefficient under strong selection (Figure \ref{fig:figure-4}B,
middle row). However, when selection was weak, the classic BGS model
erroneously estimated strong selection and a very low mutation rate. By
contrast, our B' method estimated selection coefficients much closer to their
true value. A minor discrepancy occurs around $2Ns = 1$, for reasons discussed
in Section XXX. Because we only simulated fixed selection coefficients and five
chromosomes, we only assessed the accuracy of average selection coefficients
and not the DFE probabilities.

Third, with both our B' and the classic BGS models, we find mutation rate
estimates show slight biases in the strong-selection regime (Figure
\ref{fig:figure-3}B, bottom row). Still, our B' method leads to more accurate
mutation rate estimates than classic BGS theory across a variety of selection
coefficients. When BGS is not acting, either due to weak selection or a low
rate of deleterious mutations (Figure \ref{fig:figure-4}B right column), all
estimates are worse. However, we note that this is an unlikely region of
parameter space and this issue is readily diagnosable from the poor model fit 
as evident by low $R^2$ values.
